---
title: "Fst on Ancestry"
output: html_document
date: "2024-10-29"
---

Which sites should we keep to run admixture on?
I calculated Fst on the ancestry file between samples making a pure var rudis versus samples making a pure var tuberculatus population.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(extrafont)
library(dplyr)
library(ggplot2)
library(data.table)
library(adegenet)

rm(list= ls())

#upload data
fst <- read.table("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/8.ancestry_hmm/ancestry_maf.weir.fst", sep = "\t", header = T) #fst
#remove nas from file
fst_filt <- fst[!is.na(fst$WEIR_AND_COCKERHAM_FST),]

rudis <- read.table("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/8.ancestry_hmm/var_rudis_samp.txt", sep = "\t", header = F) #var rudis samples
tub <- read.table("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/8.ancestry_hmm/var_tub_samp.txt", sep = "\t", header = F) #var tub samples

```

Plot the distribution of fst values.
Fst - measure of population differentiation, ranges from 0 to 1 with 0 being no differentiation and 1 being complete differentation. 
We are aiming for values close to 1, to get snps that are as differentiated between populations as possible. 

```{r fst_plot}
par(family = "Times New Roman")
plot(sort(fst_filt$WEIR_AND_COCKERHAM_FST), pch = 16, xlab = "site", ylab = "Fst", main = "Population differentiation per site")

t_mean <- 2*mean(fst_filt$WEIR_AND_COCKERHAM_FST) #0.296
abline(h = t_mean, col = "red")
```
How many sites do we keep as we increase the Fst threshold?

```{r loop_through_thresholds}
tvec <- quantile(fst_filt$WEIR_AND_COCKERHAM_FST, seq(0.5,1, by = 0.05))
tot <- length(fst_filt$WEIR_AND_COCKERHAM_FST)
num_sites <- matrix(NA, length(tvec))

for (t in 1:length(tvec)) {
  
  thresh <- tvec[t]
  num_sites[t] <- length(which(fst_filt$WEIR_AND_COCKERHAM_FST > thresh))
  
  
  
}

par(family = "Times New Roman")
plot(tvec, num_sites, pch = 16, xlab = "Fst threshold", ylab = "Number of sites")
abline(v = t_mean, col = "red")
tvec[which(num_sites == 2089610)]

#sites to keep - 2 times the mean
keep <- fst_filt[which(fst_filt$WEIR_AND_COCKERHAM_FST > t_mean),] #1478756 x 3
write.table(keep, "/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/8.ancestry_hmm/fst_filtered_ancestry_sites.txt", sep = "\t", col.names = T, row.names = F)

#90th quantile
tvec
keep <- fst_filt[which(fst_filt$WEIR_AND_COCKERHAM_FST > tvec[9]),] #834280 x  3
write.table(keep, "/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/8.ancestry_hmm/fst_filtered_ancestry_sites_q90.txt", sep = "\t", col.names = T, row.names = F)

#75# quantile
keep <- fst_filt[which(fst_filt$WEIR_AND_COCKERHAM_FST > tvec[6]),] #834280 x  3
write.table(keep, "/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/8.ancestry_hmm/fst_filtered_ancestry_sites_q75.txt", sep = "\t", col.names = T, row.names = F)

```


