---
title: "Compare growth rate and LD50 in control and drought conditions"
output: html_document
date: "2025-07-01"
---

Make Figure 1E panel. 

# (1) Import datasets
```{r setup, include=FALSE}

library(extrafont)
library(dplyr)
library(ggplot2)
library(car)
library(lme4)
library(lmerTest) #need to load both libraries for p values to show 
library(colourvalues)

rm(list= ls())

#upload data
setwd("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/10.EPSPS/depths")
coding <- read.table("depth.txt", sep = "\t", header = F)
EPSPS <- read.table("EPSPS.depth", sep = "\t", header = F)
metadata <- read.table("metadata_drought.txt", sep = "\t", header = T)

#Adjust col names
colnames(EPSPS) <- c("sampID", "depth")

#create new variable with samp ID
metadata$sampID <- paste(metadata$Pool, metadata$Position, sep = "_")

#calculate mean depth for the whole coding region to delete this big file from the working environment
coding_mean_depth_per_samp <- colMeans(coding[,4:283])
coding_median_depth_per_samp <- apply(coding[,4:283],2,median)
hist(coding_mean_depth_per_samp, 100)
hist(coding_median_depth_per_samp, 100)
rm(coding)
#same order as EPSPS
EPSPS$coding_genome_wide_mean <- coding_mean_depth_per_samp
EPSPS$coding_genome_wide_median <- coding_median_depth_per_samp

#normalize by genome wide per sample mean depth for coding sequences
EPSPS$depth_norm_mean <- EPSPS$depth/EPSPS$coding_genome_wide_mean
EPSPS$depth_norm_median <- EPSPS$depth/EPSPS$coding_genome_wide_median

#remove from metadata bam with problem
# "AT11_395" & "AT1_13"
idx <- c(which(metadata$sampID == "AT1_13"), which(metadata$sampID == "AT11_395"))
metadata_fil <- metadata[-idx,]

#merge metadata and depth file for treatment
full <- merge(EPSPS, metadata_fil, by = "sampID")


#ADMIXTURE
#global admixture results
k2 <- read.table("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/2.admixture/k2_structure_res.csv", 
                 sep = ",", header = T)
#Filter out outliers : P16_Nat_1_T P12_Nat_14_T
idx <- which(k2$samp == c("P16_Nat_1_T",  "P12_Nat_14_T"))
k2_fil <- k2[-c(idx),] #280 x 11
#update label to join datasets
colnames(k2_fil) <- c("X", "Label", "pair", "env", "ind", "treatment", "lat", "long", "V1", "k2.V1", "k2.V2")
fulldt <- inner_join(full, k2_fil, by="Label")


#LD50 (DROUGHT)
setwd("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/4.phenotypes")
ld50 <- read.table("ld50.txt", sep = "\t", header = T) 
fulldt$pair <- gsub("P","",as.character(fulldt$pair)) #modify column


```


We are also interested in controlling for the drought effects by testing the trade offs between normal growth and the other variables. 


```{r upload_control_and_merge}
#merge metadata and depth file for control (sibling plants)
phenos_all <- read.table("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/data/7.tradeoffs/phenos_wmeta.txt", sep = "\t", header = T)
phenos_control <- phenos_all[phenos_all$Treatment == "Control",]

#prep dataset for merging
to_dt_to_merge <- data.frame(Pair = fulldt$Pair, Ind =  fulldt$Ind, Env =  fulldt$Env, 
                             depth_norm_mean = fulldt$depth_norm_mean, Day.of.Full.Wilt =  fulldt$Day.of.Full.Wilt, 
                             Bench = fulldt$Bench, Tray = fulldt$Tray, long = fulldt$long, lat = fulldt$lat, 
                             k2.V1 = fulldt$k2.V1)
phenos_control_to_merge <- data.frame(Pair = phenos_control$Pair, Ind =  phenos_control$Ind, Env =  phenos_control$Env, 
                             growth_rate_height =  phenos_control$growth_rate_height)  
to_dt_to_merge$Pair <- gsub("P","",as.character(to_dt_to_merge$Pair)) #modify column


#merge
control <- merge(to_dt_to_merge, phenos_control_to_merge, by = c("Pair", "Ind", "Env")) #264 samples with data !

```


# (2) Drought-treated population analyses

## Calculate growth rates 

```{r growth_rate}
#calculate growth rates
#subtract height/width after drought to before drought and compare between treatments
##stem width
fulldt$growth.rate.stem<-fulldt$Stem.Width.1W.Post.Drought - fulldt$Stem.Width..mm.

fulldt <- fulldt %>% mutate(stemrate_tilldeath=fulldt$Stem.Width.at..Death.- fulldt$Stem.Width..mm.) %>%
  mutate(stemrate_combined = coalesce(stemrate_tilldeath,growth.rate.stem))
hist(fulldt$stemrate_combined,breaks=20,)

##plant height
fulldt$growth_rate_height<-(fulldt$Plant.Height.1W.Post.Drought - fulldt$Plant.Height..cm. )/7
hist(fulldt$growth_rate_height,breaks=100) #1 week growth rate, early life history

fulldt$growth_rate_height_2<-(fulldt$Plant.Height.2W - fulldt$Plant.Height.1W.Post.Drought)/7
hist(fulldt$growth_rate_height_2,breaks=100) #1 week growth rate, mid life history 

#2 values are <0 in growth rate week 1 : prpobably plant was dying or mistake while recording measure ?
#I excluce them from the dataset
tmp <- fulldt[-which(fulldt$growth_rate_height<0), ] # 276  58
fulldt <- tmp
hist(fulldt$growth_rate_height,breaks=100) #all values are now >0


fulldt$long_cols <- color_values(fulldt$long, palette = "purples")
```



## Population summaries
Do we see more signal when we focus on population-level (mean) versus individual-level data ?

```{r summarize_pop}
pop_phenos <- fulldt %>%
  group_by(Pair, Env) %>%
  summarize(
    count = n(),                   # Count of each group
    mean_growth_rate_height = mean(growth_rate_height, na.rm=T),     
    sd_growth_rate_height = sd(growth_rate_height, na.rm=T),       
    mean_growth_rate_height = mean(growth_rate_height, na.rm=T),        
    sd_growth_rate_height = sd(growth_rate_height, na.rm=T),
    mean_epsps_depth_norm = mean(depth_norm_mean, na.rm=T),        
    sd_epsps_depth_norm = sd(depth_norm_mean, na.rm=T),
    mean_day_of_full_wilt = mean(Day.of.Full.Wilt, na.rm=T),        
    sd_day_of_full_wilt = sd(Day.of.Full.Wilt, na.rm=T),
    mean_k2 = mean(k2.V1, na.rm = T),
    sd_k2 = sd(k2.V1, na.rm = T),
    long_cols = long_cols
  )


pop_phenos_drought <- pop_phenos
```


## LD50

```{r drought_pop_level}
#prepare datasets
colnames(ld50) <- c("ld50", "ld50_minus_stderr", "ld50_plus_stderr", "pair", "lat", "long", "Env"  )
ld50$pair <- as.character(ld50$pair)
pop_phenos$pair <- gsub("P","",as.character(pop_phenos$Pair))
#join dataset
ld50_pop_phenos <- inner_join(pop_phenos, ld50, by = c("pair", "Env"))

ld50_pop_phenos_drought <- ld50_pop_phenos
```




# (3) Control populations analyses

```{r growth_vs_covariates}

control_fil <- control[control$growth_rate_height >0 ,] #263 x 11
control_fil$long_cols <- color_values(control_fil$long, palette = "purples")
```



## Population-level analysis
```{r summarize_pop}
pop_phenos <- control_fil %>%
  group_by(Pair, Env) %>%
  summarize(
    count = n(),                   # Count of each group
    mean_growth_rate_height = mean(growth_rate_height, na.rm=T),     
    sd_growth_rate_height = sd(growth_rate_height, na.rm=T),       
    mean_growth_rate_height = mean(growth_rate_height, na.rm=T),        
    sd_growth_rate_height = sd(growth_rate_height, na.rm=T),
    mean_epsps_depth_norm = mean(depth_norm_mean, na.rm=T),        
    sd_epsps_depth_norm = sd(depth_norm_mean, na.rm=T),
    mean_day_of_full_wilt = mean(Day.of.Full.Wilt, na.rm=T),
    sd_day_of_full_wilt = sd(Day.of.Full.Wilt, na.rm=T),
    mean_k2 = mean(k2.V1, na.rm = T),
    sd_k2 = sd(k2.V1, na.rm = T),
    long_cols = long_cols
  )

#prepare datasets
colnames(ld50) <- c("ld50", "ld50_minus_stderr", "ld50_plus_stderr", "pair", "lat", "long", "Env")
ld50$pair <- as.character(ld50$pair)
pop_phenos$pair <- gsub("P","",as.character(pop_phenos$Pair))
#join dataset
ld50_pop_phenos <- inner_join(pop_phenos, ld50, by = c("pair", "Env"))



```
# (4) Figure 1E plot
```{r F1E}

setwd("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/figures/1.ld50_pca/")

pdf("Figure1E_stderr.pdf", 
    bg = "white", width=4, height=5, family = "Times New Roman")


par(family = "Times New Roman", mfrow = c(1,1), cex.axis = 1, cex.lab = 1.5) 

N <- length(ld50_pop_phenos_drought$ld50)

plot(1,  pch = 21, ylab = "Drought LD50", cex = 1.2,
     xlab = "Mean early growth rate", main = "", xlim = c(4,6.5), ylim = c(7,14), yaxt = "n")

#pop - control
to_c <- lm( ld50_pop_phenos$ld50 ~ ld50_pop_phenos$mean_growth_rate_height) #linear model
summary(to_c)
#correlation coefficient
cor(ld50_pop_phenos$ld50, ld50_pop_phenos$mean_growth_rate_height) #-0.5811614, p-value < 2.2e-16

cor.test(ld50_pop_phenos$ld50, ld50_pop_phenos$mean_growth_rate_height)
#se
N <- length(ld50_pop_phenos$ld50)
arrows(ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50_plus_stderr,
       ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50_minus_stderr,
       code = 3, angle = 90, length = 0, col = "#353535")

arrows(pop_phenos$mean_growth_rate_height + (pop_phenos$sd_growth_rate_height/sqrt(N)), ld50_pop_phenos$ld50, 
       pop_phenos$mean_growth_rate_height - (pop_phenos$sd_growth_rate_height/sqrt(N)), ld50_pop_phenos$ld50, code = 3,
       angle = 90, length = 0, col = "#353535")

#pch by environment
pchvec <- rep(21, length(ld50_pop_phenos$ld50))
pchvec[ld50_pop_phenos$Env == "Nat"] <- 22
points(ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50,  pch = pchvec, cex=1.2, lwd = 0.5,
      col = "#353535", bg = ld50_pop_phenos$long_cols)

#fits
points(c(seq(4.2,6.4,by = 0.2)), to_c$coefficients[1] + to_c$coefficients[2] * c(seq(4.2,6.4,by = 0.2)), type = "l", col = "#353535", lwd = 2)

dev.off()

summary(to_c)

#Wilcoxon test to compare habitats
#setp 1 : keep the same number of inds per pair
pairs <- as.data.frame(table(ld50_pop_phenos$Pair[ld50_pop_phenos$Env == "Ag"]))$Var1
ag_counts <- as.data.frame(table(ld50_pop_phenos$Pair[ld50_pop_phenos$Env == "Ag"]))$Freq
nat_counts <- as.data.frame(table(ld50_pop_phenos$Pair[ld50_pop_phenos$Env == "Nat"]))$Freq
#take the minimum of each counts
ag_mean_epsps <- c()
nat_mean_epsps <- c()
for (p in 1:length(pairs)) {
  cur_p <- pairs[p]
  min_val <- min(c(ag_counts[p], nat_counts[p]))
  #select first min_val samples
  ag_samps <- ld50_pop_phenos[ld50_pop_phenos$Pair == cur_p & ld50_pop_phenos$Env == "Ag",]
  ag_mean_epsps <- rbind(ag_mean_epsps,ag_samps[1:min_val,])
  
  nat_samps <- ld50_pop_phenos[ld50_pop_phenos$Pair == cur_p & ld50_pop_phenos$Env == "Nat",]
  nat_mean_epsps <- rbind(nat_mean_epsps,nat_samps[1:min_val,])
}

#check
dim(nat_mean_epsps) # 100 x 18
dim(ag_mean_epsps) # 100 x 18

wilcox.test(nat_mean_epsps$mean_growth_rate_height, ag_mean_epsps$mean_growth_rate_height, paired = T, alternative = "greater")
# 	Wilcoxon signed rank test with continuity correction
# 
# data:  nat_mean_epsps$mean_growth_rate_height and ag_mean_epsps$mean_growth_rate_height
# V = 3928, p-value = 6.793e-07
# alternative hypothesis: true location shift is greater than 0

```

```{r separate_habitats}

setwd("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/figures/1.ld50_pca/")

pdf("Figure1E_stderr_habitats.pdf", 
    bg = "white", width=4, height=5, family = "Times New Roman")


par(family = "Times New Roman", mfrow = c(1,1), cex.axis = 1, cex.lab = 1.5) 

N <- length(ld50_pop_phenos_drought$ld50)

plot(1,  pch = 21, ylab = "Drought LD50", cex = 1.2,
     xlab = "Mean early growth rate", main = "", xlim = c(4,6.5), ylim = c(7,14), yaxt = "n")

#pop - control + nat
idx_nat <- ld50_pop_phenos$Env == "Nat"
to_c_nat <- lm( ld50_pop_phenos$ld50[idx_nat] ~ ld50_pop_phenos$mean_growth_rate_height[idx_nat]) #linear model
summary(to_c_nat)
idx_ag <- ld50_pop_phenos$Env == "Ag"
to_c_ag <- lm( ld50_pop_phenos$ld50[idx_ag] ~ ld50_pop_phenos$mean_growth_rate_height[idx_ag]) #linear model
summary(to_c_ag)

#correlation coefficient
cor(ld50_pop_phenos$ld50, ld50_pop_phenos$mean_growth_rate_height) #-0.5811614, p-value < 2.2e-16

cor.test(ld50_pop_phenos$ld50, ld50_pop_phenos$mean_growth_rate_height)
#se
N <- length(ld50_pop_phenos$ld50)
arrows(ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50_plus_stderr,
       ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50_minus_stderr,
       code = 3, angle = 90, length = 0, col = "#353535")

arrows(pop_phenos$mean_growth_rate_height + (pop_phenos$sd_growth_rate_height/sqrt(N)), ld50_pop_phenos$ld50, 
       pop_phenos$mean_growth_rate_height - (pop_phenos$sd_growth_rate_height/sqrt(N)), ld50_pop_phenos$ld50, code = 3,
       angle = 90, length = 0, col = "#353535")

#pch by environment
pchvec <- rep(21, length(ld50_pop_phenos$ld50))
pchvec[ld50_pop_phenos$Env == "Nat"] <- 22
points(ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50,  pch = pchvec, cex=1.2, lwd = 0.5,
      col = "#353535", bg = ld50_pop_phenos$long_cols)

#fits
points(c(seq(4.2,6.4,by = 0.2)), to_c_nat$coefficients[1] + to_c_nat$coefficients[2] * c(seq(4.2,6.4,by = 0.2)), type = "l", col = "#353535", lwd = 2, lty = 2)
points(c(seq(4.2,6.4,by = 0.2)), to_c_ag$coefficients[1] + to_c_ag$coefficients[2] * c(seq(4.2,6.4,by = 0.2)), type = "l", col = "#353535", lwd = 2)

dev.off()
```



# (4) (old) Figure 1E plot 

```{r combined_all}

setwd("/Users/rozenn/Library/CloudStorage/GoogleDrive-rozennpineau@uchicago.edu/My Drive/Work/9.Science/1.DroughtProject/1.analyses/figures/1.ld50_pca/")

pdf("Figure1E.pdf", 
    bg = "white", width=4, height=5, family = "Times New Roman")


par(family = "Times New Roman", mfrow = c(1,1), cex.axis = 1.5, cex.lab = 1.5)#, cex = 1.1,  mai = c(0.6, .5, .5, .5), oma = c(.1, .1, .1, .1))

#-------------------------------------------------------------
#first row : growth_rate_height - EPSPS
line_drought <- "#62938b"
line_control <- "#4d4d4d"

N <- length(ld50_pop_phenos_drought$ld50)
ld50_pop_phenos$ld50_se <- (ld50_pop_phenos$ld50-ld50_pop_phenos$ld50_plus_stderr)/N

#pop - control
to_c <- lm( ld50_pop_phenos$ld50 ~ ld50_pop_phenos$mean_growth_rate_height)
plot(ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50,  pch = 21, ylab = "LD50", cex = 1.2,
     xlab = "Mean early growth rate", main = "", xlim = c(3.2,6), col = "gray", bg = ld50_pop_phenos$long_cols, ylim = c(7,14), yaxt = "n")
#se
N <- length(ld50_pop_phenos$ld50)
arrows(ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50 + ld50_pop_phenos$ld50_se, 
       ld50_pop_phenos$mean_growth_rate_height, ld50_pop_phenos$ld50 - ld50_pop_phenos$ld50_se, code = 3,
       angle = 90, length = 0.05, col = ld50_pop_phenos$long_cols)
arrows(pop_phenos$mean_growth_rate_height + (pop_phenos$sd_growth_rate_height/N), ld50_pop_phenos$ld50, 
       pop_phenos$mean_growth_rate_height - (pop_phenos$sd_growth_rate_height/N), ld50_pop_phenos$ld50, code = 3,
       angle = 90, length = 0.05, col = ld50_pop_phenos$long_cols)


#pop - drought
to <- lm( ld50_pop_phenos_drought$ld50 ~ ld50_pop_phenos_drought$mean_growth_rate_height)
points(ld50_pop_phenos_drought$mean_growth_rate_height, ld50_pop_phenos_drought$ld50,  pch = 8, bg = ld50_pop_phenos$long_cols, col = "gray", cex = 1.2)
#se
N <- length(ld50_pop_phenos_drought$ld50)
ld50_pop_phenos_drought$ld50_se <- (ld50_pop_phenos_drought$ld50_plus_stderr-ld50_pop_phenos_drought$ld50)/N
arrows(ld50_pop_phenos_drought$mean_growth_rate_height, ld50_pop_phenos_drought$ld50 + ld50_pop_phenos_drought$ld50_se, 
       ld50_pop_phenos_drought$mean_growth_rate_height, ld50_pop_phenos_drought$ld50 - ld50_pop_phenos_drought$ld50_se, 
       code = 3, angle = 90, length = 0.05, col = ld50_pop_phenos_drought$long_cols)
arrows(ld50_pop_phenos_drought$mean_growth_rate_height + ld50_pop_phenos_drought$sd_growth_rate_height/N, ld50_pop_phenos_drought$ld50, 
       ld50_pop_phenos_drought$mean_growth_rate_height - ld50_pop_phenos_drought$sd_growth_rate_height/N, ld50_pop_phenos_drought$ld50, 
       code = 3, angle = 90, length = 0.05, col = ld50_pop_phenos_drought$long_cols)
#fits
points(c(seq(4,6,by = 0.2)), to_c$coefficients[1] + to_c$coefficients[2] * c(seq(4,6,by = 0.2)), type = "l", col = line_control, lwd = 2)
points(c(seq(3.2,6,by = 0.2)), to$coefficients[1] + to$coefficients[2] * c(seq(3.2,6,by = 0.2)), type = "l", col = line_drought, lwd = 2)

dev.off()


```





